\batcave: Bayesian Analysis Tools for Context-Aware Variant Evaluation

### 2019-09-24 NAR_DRAFT.tex

\batcave divides the data into two classes: high- and low-confidence variants.
The high-confidence variants are used to estimate the mutational prior, which is then used to improve the calling of low-confidence variants.
Statistically, this is an empirical Bayesian approach~\citep{Robbins1954}, in which the high and low-confidence variants are treated as parallel experiments\citep{Morris1983,Efron2014}. 
In general, high-confidence variants tend to have relatively high allelic frequencies, and consequently tend to have arisen early in tumor development.
A key implicit assumption of our approach is that the mutational process does not change between high- and low-confidence variants, implying that the mutational profile of the tumor is temporally constant.
Recent studies have found differences in mutational profiles among variants of different allelic frequencies~\citep{Rubanova2018a}, although those differences are relatively small.
A potential extension of the \batcave algorithm is process potential variants in order of descending allelic frequency and to update the estimated mutational prior as the algorithm proceeds.
This approach might increasing sensitivity to low-frequency variants generated by recently-arisen mutational processes, at the cost of potentially increasing sensitivity to patterns of sequencing error.

%\batcave resembles an empirical Bayes approach, but is not really empirical Bayes.
%In fact, variants are best described as sequential samples from a data generating distribution represented by the mutation profile.
%Early mutations are classified with low error, while later mutations are classified with less certainty.
%We show here that sequential updating of the dirichlet-multinomial posterior converges to the true data generating distribution using the set of early mutations.
%Along with the assumption that tumor evolution is approximately neutral, this allows us to generate a tumor and site-specific prior that sharpens the classification and calibration of later mutations.
%Thus, \batcave is best described as an algorithm trained on early mutations.

We evaluated \batcave with simulated tumors with three different mutation profiles and two real tumors.
The simulated diffuse and intermediate profiles (Fig.~\ref{NAR-simsigfig}A\&B) represent baseline profiles of lung and breast tumors, respectively, which the concentrated profile (Fig.~\ref{NAR-simsigfig}C) represents a tumor driven by a particular mutational process, such as UV exposure.
But mutational profiles are highly heterogeneous, so that concentrated profiles can be found in any tumor type (e.g., Fig.~\ref{NAR-realsigfig}C).
The two real data sets we considered are among the very few for which extensive validation of variant calls has been performed~\cite{Griffith2015, Shi2018}.
They happen, however, to have relatively flat mutation profiles (Fig.~\ref{NAR-realsigfig}A\&B), which reduces the expected advantage of \batcave over MuTect (Table~\ref{table:01}).
A more fundamental challenge of using these real data for testing callers is that only a subset of potential variants are validated.
This subset tends to be relatively easy to call, so both MuTect and \batcave have almost perfect precision and recall (Fig.~\ref{NAR-wes_fig} and Table~\ref{table:01}).
Moreover, few true negative sites are validated, so specificity is impossible to calculate. \rngcomment{Double-check my logic here...}

@@

>>>>

\batcave divides the data into two classes: high- and low-confidence variants.
The high-confidence variants are used to estimate the mutational prior, which is then used to improve the calling of low-confidence variants.
Statistically, this is an empirical Bayesian approach~\citep{Robbins1954}, in which the high and low-confidence variants are treated as parallel experiments \citep{Morris1983,Efron2014}. 
In general, high-confidence variants tend to have relatively high allelic frequencies, and consequently tend to have arisen early in tumor development.
A key implicit assumption of our approach is that the mutational process does not change between high- and low-confidence variants, implying that the mutational profile of the tumor is temporally constant.
Recent studies have found differences in mutational profiles among variants of different allelic frequencies~\citep{Rubanova2018a}, although those differences are relatively small.
A potential extension of the \batcave algorithm is process potential variants in order of descending allelic frequency and to update the estimated mutational prior as the algorithm proceeds.
This approach might increase sensitivity to low-frequency variants generated by recently-arisen mutational processes, at the cost of potentially increasing sensitivity to patterns of sequencing error.

We evaluated \batcave with simulated tumors with three different mutation profiles and two real tumors.
The simulated diffuse and intermediate profiles (Fig.~\ref{NAR-simsigfig}A\&B) represent baseline profiles of lung and breast tumors, respectively, which the concentrated profile (Fig.~\ref{NAR-simsigfig}C) represents a tumor driven by a particular mutational process, such as UV exposure.
But mutational profiles are highly heterogeneous, so that concentrated profiles can be found in any tumor type (e.g., Fig.~\ref{NAR-realsigfig}C).
The two real data sets we considered are among the very few for which extensive validation of variant calls has been performed~\cite{Griffith2015, Shi2018}.
They happen, however, to have relatively flat mutation profiles (Fig.~\ref{NAR-realsigfig}A\&B), which reduces the expected advantage of \batcave over MuTect (Table~\ref{table:01}).
A more fundamental challenge of using these real data for testing callers is that only a subset of potential variants are validated.
This subset tends to be relatively easy to call, so both MuTect and \batcave have almost perfect precision and recall (Fig.~\ref{NAR-wes_fig} and Table~\ref{table:01}).
Moreover, few true negative sites are validated, so specificity is impossible to calculate.

@@

We focus on the calibration benefits of \batcave because calibration is important for both clinical and research communities.
\begin{itemize}
  \item Research/Discovery
  \begin{itemize}
    \item Manual review is time consuming, and the number of mutations can be overwhelming at lower frequencies. Calibration reduces wasted effort.
    \item Drug discovery is expensive and effort must be focused. Finding common low frequency alleles is important, and having a calibrated probability for each tumor a variant is found in would be useful.
  \end{itemize}
  \item clinical
  \begin{itemize}
    \item many resistance mutations are present at low frequencies in primary tumor. If a tumor is sequenced and you can assign a probability to the mutation, it may be possible to indicate treatment with the drug for the resistance mutation simultaneous to the first line treatment. For instance if the mutation has a 80\% probabiliity of being present, yes, but if 20\% no. Balance of harm from treating for a mutation that isn't there.
  \end{itemize}

\end{itemize}
\bkmcomment{begin needed paragraph above}
